---
title: "Ratios"
output: html_notebook
---


# Goals

- Practice plotting with ggplot.
- Practice building more simple and multiple linear regressions.
- This week, YOU have to create your own code chunks.
    - I recommend separating your plot code chunks from your modeling code chunks to help you see everything.


# Setup

```{r message=FALSE, warning=FALSE}
library(janitor)
library(lubridate)
library(modelr)
library(tidyverse)
library(Stat2Data)
data(Hawks)

# This is what we would call a data pipe-line. It makes the data easier to work with.
hawks <-
  Hawks |>
  # Lower case column names. And Spaces rather than SnakeCase.
  clean_names() |>
  # Adds some actual dates data and makes species easier to read.
  mutate(
    capture_date = ymd(paste(year, month, day)),
    capture_day_of_week = wday(capture_date, label = TRUE),
    species_desc = case_when(
      species == "SS" ~ "Sharp-Shinned",
      species == "RT" ~ "Red-Tailed",
      species == "CH" ~ "Cooper's"
    )
  ) |>
  # Rearranges our columns.
  select(
    band_number,
    capture_date,
    capture_time,
    capture_year = year,
    capture_month = month,
    capture_day = day,
    capture_day_of_week,
    species,
    species_desc,
    age:crop
  )
rm(Hawks)
```



# Data - Hawks (again)

## Data

### Description:

Data for a samples of hawks from three different species

### Format:

A data frame with 908 observations on the following 19 variables.

- band_number:	ID band code
- capture_date:	Date the bird was captured
- capture_time:	Time of capture (HH:MM)
- capture_year: 1992-2003 - The year the bird was captured
- capture_month: 9=September to 12=December - The month the bird was captured
- capture_day: The day the bird was captured
- capture_day_of_week: The name of the day of the week the bird was captured
- species: CH=Cooper's, RT=Red-Tailed, SS=Sharp-Shinned
- age: A=Adult or I=Immature
- sex: F=Female or M=Male
- wing: Length (in mm) of primary wing feather from tip to wrist it attaches to
- weight: Body weight (in gm)
- culmen: Length (in mm) of the upper bill from the tip to where it bumps - into the fleshy part of the bird
- hallux: Length (in mm) of the killing talon
- tail: Measurement (in mm) related to the length of the tail (invented at the MacBride Raptor Center)
- standard_tail: Standard measurement of tail length (in mm)
- tarsus: Length of the basic foot bone (in mm)
- wing_pit_fat: Amount of fat in the wing pit
- keel_fat: Amount of fat on the breastbone (measured by feel
- crop: Amount of material in the crop, coded from 1=full to 0=empty

### Details:

Students and faculty at Cornell College in Mount Vernon, Iowa, collected data over many years at the hawk blind at Lake MacBride near Iowa City, Iowa. The data set that we are analyzing here is a subset of the original data set, using only those species for which there were more than 10 observations. Data were collected on random samples of three different species of hawks: Red-tailed, Sharp-shinned, and Cooper's hawks.

### Source

Many thanks to the late Professor Bob Black at Cornell College for sharing these data with the world.


# Lab

- Because I'm a bird brain, we shall continue to focus on body mass of birds.
- Because why not?
- Be sure to PLOT and MODEL each of these as appropriate.
  - A model without a plot is a dangerous thing.

## Task 01

I'd like to better understand the average body weight difference between an immature and adult For the Red-Tailed hawk. Build a linear regression model that models weight as a function of age (Immature/Adult) for the Red-Tailed hawk.

- **Question:** According to this model, which is heavier, immature or adult hawks and are the results statistically significant?
  - Build a model called lm_age.
  - Plot this using ggplot using the box-plot geom.
- **Answer:** 

```{r}
ggplot(hawks, aes(x = age, y = weight)) +
  geom_boxplot()
```

```{r}
lm_age <- lm(weight ~ age, data = hawks)
summary(lm_age)
```

## Task 02

Do these results surprise you? Starting with your plot from Task 01, can you dig into this and come up with an explanation for what is causing this to happen in our regression?

- **Question:** What feature from our data best explains why immature hawks appear to be slightly heavier than adult hawks?
  - Hint: The answer is a categorical variable!
- **Answer:** 

```{r}
ggplot(hawks, aes(x = age, y = weight)) +
  geom_boxplot() +
  facet_wrap(~species)

ggplot(hawks, aes(x = age)) +
  geom_bar() +
  facet_wrap(~species)

ggplot(hawks, aes(x = age, y = weight)) +
  geom_boxplot() +
  facet_wrap(~sex)

ggplot(hawks, aes(x = age)) +
  geom_bar() +
  facet_wrap(~sex)
```


## Task 03

I would like to know if adult Red-Tailed hawks got heavier, lighter, or stayed the same during the collection period.

- **Question:** Did adult birds get heavier, lighter, or mostly stay the same during the collection period? Are the results statistically significant?
  - For the plot, put capture_date on the X Axis and weight on the Y Axis.
  - Watch out - we are only looking at ADULT RED-TAILED hawks. Exclude all other hawks from this analysis.
  - Build a linear regression to tell us about the slope of any observed change and if the change is statistically significant.
    - Name the model lm_capture_date.
    - Is the model statistically significant?
    - What do you think this means in this context?
- **Answer:** 

```{r}
ggplot(
  data = hawks |> filter(age == "A", species == "RT"),
  aes(x = capture_date, y = weight)
) +
  geom_point() +
  geom_smooth(method = "lm")

lm_capture_date <- lm(
  weight ~ capture_date,
  data = hawks |> filter(age == "A", species == "RT")
)
summary(lm_capture_date)
```


## Task 04:

- **Question:** Which is a better predictor of hawk body mass?
  - Wing length (wing)
  - Tail length (tail)
  - In Canvas, I'm going to ask you to write a few sentences supporting WHY you chose wing or tail.
- **Answer:** 

As always, I recommend graphing this AND building the model. For this task, build two models:

- lm_wing
- lm_tail

```{r}
ggplot(hawks, aes(x = wing, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm")

lm_wing <- lm(weight ~ wing, data = hawks)
summary(lm_wing)
```

```{r}
ggplot(hawks, aes(x = tail, y = weight)) +
  geom_point() +
  geom_smooth(method = "lm")

lm_tail <- lm(weight ~ tail, data = hawks)
summary(lm_tail)
```


## Task 05:

- **Question:** Plot the distribution of the residuals of lm_wing using geom_density and geom_qq. Do these residual values appear to be normally distributed?
- **Answer:** 

```{r}
ggplot(hawks |> add_residuals(lm_wing), aes(resid)) +
  geom_density()

ggplot(hawks |> add_residuals(lm_wing), aes(sample = resid)) +
  geom_qq() +
  geom_qq_line()
```

## Task 06:

- **Question:** Plot residuals as a function of predicted values. Do our errors look homogenous?
- **Answer:** 

```{r}
ggplot(
  data = hawks |> add_residuals(lm_wing) |> add_predictions(lm_wing),
  aes(x = pred, y = resid, color = species)
) +
  geom_point() +
  geom_smooth()
```

## Task 07:

- **Question:** Add species to lm_wing and tell me if you think we should add it to the model with or without interactions. In other words, do we need to adjust the slope, change the y-intercept, or both for each species?
  - Graph this using ggplot and use that to inform your choice.
  - Build the model and tell me what the Adjusted R-Squared value is.
  - Call your new model lm_wing_species.
- **Answer:** 

```{r}
ggplot(hawks, aes(x = wing, y = weight, color = species)) +
  geom_point() +
  geom_smooth(method = "lm")

lm_wing_species <- lm(weight ~ wing, data = hawks)
summary(lm_wing)
```

But we could do this.

```{r}
hawks |>
  group_by(species) |>
  summarize(
    avg_weight = mean(weight, na.rm = TRUE),
    sd_weight = sd(weight, na.rm = TRUE),
    avg_wing = mean(wing, na.rm = TRUE),
    sd_wing = sd(wing, na.rm = TRUE)
  ) |>
  mutate(
    min_weight = avg_weight - 2 * sd_weight,
    max_weight = avg_weight + 2 * sd_weight,
    min_wing = avg_wing - 2 * sd_wing,
    max_wing = avg_wing + 2 * sd_wing
  ) |>
  select(species, min_weight, max_weight, min_wing, max_wing)

hawks <-
  hawks |>
  mutate(
    not_extreme = case_when(
      species == "CH" & (weight < 96.4 | weight > 745) ~ FALSE,
      species == "CH" & (wing < 180 | wing > 308) ~ FALSE,
      species == "RT" & (weight < 716 | weight > 1473) ~ FALSE,
      species == "RT" & (wing < 320 | wing > 446) ~ FALSE,
      species == "SS" & (weight < -13.3 | weight > 309) ~ FALSE,
      species == "SS" & (wing < 140 | wing > 230) ~ FALSE,
      TRUE ~ TRUE
    )
  )

hawks |>
  filter(!not_extreme) |>
  count(species)

ggplot(
  hawks |> filter(not_extreme),
  aes(x = wing, y = weight, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")

lm_wing_species_no_outliers <-
  lm(weight ~ wing * species, data = hawks |> filter(!not_extreme))
summary(lm_wing_species_no_outliers)

hawks |>
  filter(not_extreme) |>
  add_residuals(lm_wing_species_no_outliers) |>
  ggplot(aes(sample = resid)) +
  geom_qq() +
  geom_qq_line()
```

## Task 08:

The following questions are based on the plot you created for Task 07. This plot should be weight as a function of wing length with species colored in. 

- **Question:** Given a hypothetical hawk of weight 400 and wing length 1,000. Guess which species this hawk is most likely to be. How confident are you that your guess is correct? (High, Medium, Low) 
- **Answer:** 

- **Question:** Given a hypothetical hawk of weight 200 and wing length 200. Guess which species this hawk is most likely to be. How confident are you that your guess is correct? (High, Medium, Low) 
- **Answer:** 
