---
title: "Ratios"
output: html_notebook
---


# Goals

- Learn about Risk Ratio
- Learn about Odds Ratio
- Learn how to calculate these metrics using the tools we have learned in class.
- Connection to Logistic regression



# Setup

```{r message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
rm(list = ls())
library(broom)
library(tidymodels)
library(tidyverse)
strokes <- read_csv("data/strokes.csv")
```



# Data - Strokes

Source: [Kaggle Stroke Prediction Dataset](https://www.kaggle.com/fedesoriano/stroke-prediction-dataset)

> According to the World Health Organization (WHO) stroke is the 2nd leading cause of death globally, responsible for approximately 11% of total deaths. This dataset is used to predict whether a patient is likely to get stroke based on the input parameters like gender, age, various diseases, and smoking status. Each row in the data provides relavant information about the patient.

```{r paged.print=TRUE}
strokes
```

1. id: unique patient identifier
2. gender: "Male", "Female" or "Other"
3. age: age of the patient
4. hypertension: 0 if the patient doesn't have hypertension, 1 if the patient has hypertension
5. heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the patient has a heart disease
6. ever_married: "No" or "Yes"
7. work_type: "children", "Govt_jov", "Never_worked", "Private" or "Self-employed"
8. residence_type: "Rural" or "Urban"
9. avg_glucose_level: average glucose level in blood
10. bmi: body mass index
11. smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12. stroke: 1 if the patient had a stroke or 0 if not
*Note: "Unknown" in smoking_status means that the information is unavailable for this patient



# Lab

## Task 01

- Question: The data includes some information about the employment of the patients. Which work type is the most prevalent in this data set?
- Answer:To prevent confusion, your options below are identical to what you should have found in your data. In Canvas complete this task with one of the following:

    1. children
    2. Govt_job
    3. Never_worked
    4. Private
    5. Self-employed

Helpful Hints: The entries in the `work_type` column are formatted oddly, but that's OK. There's always _something_ funny about a data set.

```{r}
## YOUR CODE HERE!!!
```

## Task 02

- Question: Use R to calculate the percent of patients who are children (based on work type/Task 01).
- Answer: 

Helpful Hints:

- In other words, divide the work_type count by the total.
- And yes, you CAN do this without R, but the point is to do this using R.
- And the easiest way is to calculate the percent for all of the work types.

```{r}
## YOUR CODE HERE!!!
```


## Task 03

- Question: Which patients are more likely to have had a stroke, men or women? 
- Answer:

Helpful Hints:

- We want to calculate the percentage of patients with a stroke.
    - Call the number of patients with a stroke `diseased`.
    - Call the number of total patients `total`.
    - You can calculate the percent (risk) using these columns in your summarize function. Or, you can add a mutate command and do it there. The answer will be the same.
- And we want that answer for men and women.
    - group_by MUST come before summarize
    - Unless you want to rename the column, you do not need ANY equal signs in the group_by parameters.

```{r}
## YOUR CODE HERE!
```

We would thus say that men have a HIGHER risk of stroke than women, based on this data.


## Task 04

In Task 03, we saw that men have an over-all higher risk of stroke than women. Do men with hypertension have a higher risk of stroke than women with hypertension?

- This is ALMOST the same question.
- All we need to do is to filter down to those with hypertension.
- Generally, filter should come before group by. Not always, but it's a nice rule-of-thumb.
- Remember, filter needs at least one Boolean parameter.

```{r}
## YOUR CODE HERE!!!
```


# Lab - Risk Ratio


We will now introduce a new idea called the Risk Ratio and we will demonstrate how this ratio connects to the skills we have learned in this class. Our goal is to compare the risk of stroke in patients 65 and older to the risk of stroke in younger patients.

- You should know, by now, that older patients are more likely to have a stroke.
- But we don't have a TRUE/FALSE column for older patients, so we have to make it.
    - This is similar to what we did above, but this time we want to create a new column in our strokes data set.
    - And we want this in our source data, because we are going to spend a lot of time discussing this.

## Task 05

- Question: Which age group has a higher risk for stroke?
    - Under 65
    - 65 and older
- Answer: 

Helpful Hints:

- Create a new data set called RiskStroke as part of this question.
- Create a new column which identifies patients who are under/over age 65.
- Then calculate the risk of stroke for each group.


```{r}
## YOUR CODE HERE!!!
## Remember to create a new table called RiskStroke.
```

## Task 06

- Question: What is the risk ratio for older adults compared to younger patients?
- Answer: 


```{r}
## YOUR CODE HERE!
```

## Task 07

- Question: What is the odds ratio for older adults compared to younger patients for stroke?
- Answer: 

```{r}
## YOUR CODE HERE!!!
## Create an object called OddsStroke
```

## Task 08

Demonstrate that the same odds ratio can be obtained via logistic regression.

```{r}
## YOUR CODE HERE!!!
```

Probability ranges from negative infinity to positive infinity. And, we can't model infinity. So a logistic regression model models the log of this because this constrains the values of the number between 0 and 1 but it means we have to take the exponent of the estimate.

For more on this topic: [There's Nothing Odd About Th e Odds Ratio](https://www.statisticssolutions.com/theres-nothing-odd-about-the-odds-ratio-interpreting-binary-logistic-regression/)

Logistic regression allows us to to model more complex models and relationships than we would ever want to model by hand.

