---
title: "Simple & Multiple Linear Regression"
format: html 
---



Goals
================================================================================

- Recap: Simple Linear Regression (SLR)
    - Detailed example (insurance charges)
    - Assumptions/rules
- Introduce: Multiple Linear Regression (MLR)
    - Formula
    - Example: extending our model from SLR
    - Assumptions/rules


```{r}
#| label: setup
#| include: false

library(gt)
library(patchwork)
library(tidymodels)
library(tidyverse)

insurance <- read_csv("data/insurance.csv")

# Mostly disables scientific notation
options(scipen = 999)

```



Simple Linear Regression
================================================================================

Last week, we introduced Simple Linear Regression. This just means we modeled a
variable y as a function of a variable x. For example, earlier in the semester,
you created the following plot.


```{r}

ggplot(insurance, aes(x = age, y = charges)) +
  geom_point() +
  geom_smooth(method = lm)

```

- At the time I asked you if there appeared to be a relationship between age and
charges. Hopefully, you answered affirmatively.
- Last week, we learned how to calculate the regression line shown in the above
  plot.
- This week, we will go further and try to better understand what makes SLR so
  powerful.
- First, we will create a model, called `lm_simple` which models insurance
  charges as a function of age.
    - We can see it, but ggplot doesn't give us all the details.


```{r}

# Build the model
lm_simple <-
  linear_reg() |>
  fit(charges ~ age, data = insurance)

# See the model coefficients (rounded)
lm_simple

# See the model details.
glance(lm_simple)

```

- On average, people get $258 more expensive for year aged.
- Although statistically significant, this model only explains 9% of the
  variance in cost.
      - This is not actually a very good model.
- Of course, you already knew that, just by looking at the first plot.
- One important way to evaluate a model is to look at the residuals.
    - Residuals are the difference between the actual value and the predicted
      value.
- The following code adds the residuals from `lm_simple` to the insurance data.
    - Tidymodels can "augment" our data, by adding predicted values `.pred` and
      residuals `.resid` to our data.


```{r}

# By default, .pred and .resid become the first two columns.
# I am moving them to the end to make it easier to read.
insurance <-
  lm_simple |>
  augment(insurance) |>
  select(
    age,
    sex,
    bmi,
    children,
    smoker,
    region,
    charges,
    # Changing these column names will make sense later.
    pred_simple = .pred,
    resid_simple = .resid
  ) 

# We will use the first TWO rows of our data to better understand this.
insurance |> head(2)

```


Unpacking SLR
--------------------------------------------------------------------------------

- **Question:** Why does the model predict $8,063 for the first row?
- **Answer:** Because, linear regression.

$$y = a \cdot x + b$$

- `y` is the predicted insurance cost ($8,062.614)
- `a` is the the slope of the model, the increase in member charges for every
  year older ($257.7226)
- `x` is the age of the member (19)
- `b` is the intercept value ($3,165.8850)
- Note: The values shown when you call the model are rounded for clarity.
    - To access more exact estimates: `lm_simple$fit$coefficients`
    - You don't need to know how to do this, but I wanted the numbers to work.

Pull out your phone and do the arithmetic with me.

$$257.7226 \cdot 19 + 3,165.8850$$

- Is there some rounding going on? Yes. Does this work? yes
- But . . . Residuals.
- **Row 1:** A 19 year old with nearly $17,000 in charges.
    - Predicted cost: $8,063
    - `16885 - 8063 = 8822`
- **Row 2:** An 18 year old with about $1,700 in charges.
    - Predicted cost: $7,805
    - `1726-7805 = -6079`
- The difference in the predicted cost between our 18 year old and 19 year
  old?
    - `8063 - 8705 = 258`
- There are (many) other factors affecting charges which are not accounted for
  in this model, but the model is internally consistent.
      - But it isn't a very good model.
      - Look, number of people with similar actual costs is basically . . . . nobody.
- One way to evaluate a SLR is to look at the average and standard deviation of
  the residuals.
    - Visually, this looks like the observed values sitting close to the
      predicted values (regression line).
    - Although our AVERAGE residual is small, the standard deviation is more
      than $11,000.
- We also want our residuals to be normally distributed.
    - Which they are not. At all.
- See below:


```{r}

tbl_resid <- 
  insurance |>
  summarize(
    avg_residual = format(mean(resid_simple), scientific = F),
    sd_residual = sd(resid_simple)
  )

plt_resid <-
  insurance |>
  ggplot(aes(x = resid_simple)) +
  geom_density(color = "darkred") +
  xlab("Model Residuals")

plt_resid / wrap_table(tbl_resid)

```

- We've already talked about this.
- Last time we looked at this data, we discussed the prominent banding in the
  scatter plot.


```{r}

ggplot(insurance, aes(x = age, y = charges)) +
  geom_point() +
  geom_smooth(method = lm)

```

- At the time, I asked you to tell me which variable best "explained" this
  banding:
    - `sex`
    - `smoker`
    - `region`
- I asked you to plot the data three times to determine which feature did the
  best job separating the bands.

```{r}

plt_sex <-
  ggplot(insurance, aes(x = age, y = charges, color = sex)) +
  geom_point() +
  geom_smooth(method = lm)

plt_smoker <-
  ggplot(insurance, aes(x = age, y = charges, color = smoker)) +
  geom_point() +
  geom_smooth(method = lm)

plt_region <-
  ggplot(insurance, aes(x = age, y = charges, color = region)) +
  geom_point() +
  geom_smooth(method = lm)

(plt_sex / plt_smoker / plt_region)

```

Of these three options, `smoker`, does the best job explaining the three bands.



```{r}

plt_smoker

```


Assumptions/Rules
--------------------------------------------------------------------------------

- **Continuous Dependent Variable:** The dependent variable must be
  continuous.
- **Linearity:** The relationship between the dependent and independent
  variable must be linear.
    - Apply the "eye-ball" test.
- **Independence of errors:** There is not a relationship between the residuals
and the dependent variable; in other words, residual is independent of errors.
- **Normality of errors:** The residuals must be approximately normally
  distributed.
    - Use a density plot/histogram.
- **Equal variances:** The variance of the residuals is the same for all values
  of the independent variable.
    - Avoid Bow-tie, megaphone variances.
- **No Significant Outliers:** The model cannot be driven by outlier or
  highly influential points.


Multiple Linear Regression
================================================================================

We have already defined Simple Linear Regression as:

$$y = a \cdot x + b$$

Multiple Linear Regression takes this one step farther and says that for any
arbitrary number of independent variables, `x`, we define `y` as:

$$y = a_1 \cdot x_1 + b_1 + a_2 \cdot x_2 + b_2 + a_i \cdot x_i + b_1 $$

- There is no technical limit to the number of features we can add to our model.
- That said, the goal is to create a generalized model which will work beyond
  the data used to train (fit) the model.
    - Each additional independent variable makes understanding the model more
      complex.
    - Adding too many variables is called over-fitting the model.
    - This can cause the model to work well for the data used to train the
      model, but not work well with new cases.
- To add `smoker` to our simple linear regression, we can simply add the
  variable to our model formula.
- The following code chunk creates a new model, `lm_multiple`, which uses both
  age and smoking status to explain insurance charges.


```{r}

lm_multiple <-
  linear_reg() |>
  fit(formula = charges ~ age + smoker, data = insurance)
lm_multiple
glance(lm_multiple)

```

Once we have our new model, we can use `augment()` to add our predicted values
back into insurance and to review the impact on our residuals.



```{r}

# By default, .pred and .resid become the first two columns.
# I am moving them to the end to make it easier to read.
insurance <-
  lm_multiple |>
  augment(insurance) |>
  select(
    age,
    sex,
    bmi,
    children,
    smoker,
    region,
    charges,
    # NOW I hope this makes sense!!
    pred_simple,
    resid_simple,
    pred_multiple = .pred,
    resid_multiple = .resid
  ) 

# We will use the first TWO rows of our data to better understand this.
insurance |> head(2)

```

For our 19 year old smoker, our equation becomes:

$$274.9 \cdot 19 + $2,3885.5 + -2391.6$$

While for our 18 year old non-smoker, our equation is:

$$274.9 \cdot 18 + 0 + -2391.6$$

Please note that while the predicted charges for these two cases are quite
different, the only information used to predict charges are age and smoking
status. Sex, BMI, Region, etc. are not used to predict charges. You could make a
more complex model, but we won't do so during this lecture.

Because I want to review how this more complex model impacts the over-all
distribution of our residuals:


```{r}

tbl_resid <- 
  insurance |> summarize(
  avg_residual_simple = format(mean(resid_simple), scientific = F),
  sd_residual_simple = sd(resid_simple),
  avg_residual_multiple = format(mean(resid_multiple), scientific = F),
  sd_residual_multiple = sd(resid_multiple)
  )

plt_resid <-
  insurance |>
  ggplot(aes(x = resid_simple)) +
  geom_density(color = "darkred") +
  geom_density(mapping = aes(x = resid_multiple), color = "darkblue") +
  xlab("Model Residuals")

plt_resid / wrap_table(tbl_resid)

```

- In both `lm_simple` and `lm_multiple`, the average residual is small.
- This is because it is the total AVERAGE of all the residuals.
- But the standard deviation went from $11,556 to $6,392.
- And you can see in the density plot that the residuals come closer to being
  normally distributed.

**Both simple and multiple linear regressions are powerful tools.**


Assumptions/rules
--------------------------------------------------------------------------------

- All rules from SLR carry-over to MLR.
- **Multiple Independent Variables:** Must have two or more independent
  variables.
    - These can be continuous or categorical.
- **Multicollinearity:** Independent variables may not be highly correlated with
  each other.
