---
title: "More Linear Regression"
---

```{r}
library(modelr)
library(rio)
library(tidyverse)

brfss <- import("data/brfss.csv")
penguins <- import("data/penguins.csv")
diamonds <- import("data/diamonds.csv")
```


# Simple Linear Regression

Predicting penguin body mass as a function of flipper length.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm")
```

- Yes, penguins with longer flippers also weigh more, on average.
- This appears to be a nice linear relationship.
- And we learned how to model this.

```{r}
lm_flipper <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
summary(lm_flipper)
```

- This model explains nearly 76% of the variance in body mass.
- The p-value for both our feature (flipper length) and the model are statistically significant, so we accept the model.

What we didn't talk about BEFORE spring break is how to assess your model.

- Residuals should be normally distributed.
- Homogeneity of variance.
- Error terms should be independent.

Remember that the residuals are the difference between the difference between the predicted and actual value of y for a given x. And these residuals should be normally distributed.
```{r}
penguins |>
  add_residuals(lm_flipper) |>
  ggplot(aes(resid)) +
  geom_density()
```

And while that works, a more sensitive visualization is the Q-Q plot.

- Q-Q stands for Quantile-Quantile plot.
- This is a graphical test to see if our data, residuals, come from a normal distribution.

```{r}
penguins |>
  add_residuals(lm_flipper) |>
  filter(!is.na(resid)) |>
  ggplot(aes(sample = resid)) +
  geom_qq() +
  geom_qq_line()
  ```

- And the answer here is yes, because our residuals mostly fall along the 45 degree slope.

We also want to see to test for homogeneity of variance (fancy term, simple idea).

```{r}
penguins |>
  add_residuals(lm_flipper) |>
  add_predictions(lm_flipper) |>
  filter(!is.na(resid)) |>
  ggplot(aes(x = pred, y = resid)) +
  geom_point() +
  geom_smooth()
```

- You basically want to see random noise.
- And we do, mostly.
- It does rather look like there are a couple of distinct groups.
  - Let's graph this with sex and species.

Conclusion: Although this model is good, it could be better. And to make it better, we need multiple linear regression.

# Multiple Linear Regression

Multiple linear regression is a regression model that estimates the relationship between a quantitative dependent variable (body mass) and two or more independent variables using a straight line.

But as we shall see, there are two different ways this can play out.

```{r}
## There is one penguin for which we do not have a sex recorded.
## I am filtering that single penguin out because it ruins the plot.
ggplot(penguins |> filter(!is.na(sex)), aes(x = flipper_length_mm, y = body_mass_g, color = sex)) +
  geom_point() +
  geom_smooth(method = "lm")
```

- This certainly shows us that female penguins, on average, weigh less than male penguins for a given flipper length.
- I'm saying "on average" a lot. And this isn't just a phrase.
- You can find some female penguins that are heavier than some male penguins.
- But on average, the opposite is true.

```{r}
lm_flipper_sex <- lm(
  body_mass_g ~ flipper_length_mm + sex,
  penguins |> filter(!is.na(sex))
)
summary(lm_flipper_sex)
```

- Our model now explains ~80% of the variance in body mass.
- So this slightly more complex model _is_ better.
- But not a lot.
- Let's discuss our y-intercept and slope.
- Our plot, above, suggests that slope should be about the same, but that our y-intercept differs by gender.
- And we can see that in our model.

Let's calculate the body mass of an average male/female penguin with a flipper length of 200 mm.

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  geom_smooth(method = "lm")
```

- Now let's look at species.
- The slope of the Adelie and Chinstrap are very similar, but they have slightly different y intercepts.
- The slope of Gentoos is completely different.
- When we have different slopes, we have to use **** linear regression.

```{r }
# Note: Last time, I ADDED sex because all I wanted was a different y-intercept.
#       This time I MULTIPLY by species, because I want to show an interaction.
#       This will give me both a different y-intercept and a different slope.
lm_flipper_species <- lm(body_mass_g ~ flipper_length_mm * species, penguins)
summary(lm_flipper_species)
```

- This looks a lot more complex.
  - Because it gives us both a different SLOPE and Y-INTERCEPT.
- Let's find the body mass of three penguins with 200 mm long flippers.

```{r}

```

- Note that this model explains a little bit less of the variance.
  - Which model is better?
  - Well, it depends!
  - There are not simple questions in this class, sorry.

